openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-05-24 2023-05-revision-1/main.tex
paper: clarify what prepare_partial does

We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
Next, we import the \opentsne library:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
\end{CodeInput}
\end{CodeChunk}
We now mirror the three t-SNE steps from the previous section. First, we select an affinity model which describes the similarities between data points:
\begin{CodeChunk}
\begin{CodeInput}
>>> affinities = openTSNE.affinity.PerplexityBasedNN(
...     iris.data, perplexity=30
... )
\end{CodeInput}
\end{CodeChunk}
Next, we determine the initial point positions in the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> initialization = openTSNE.initialization.pca(iris.data)
>>> initialization[:3]
array([[-1.30971087e-04,  1.55848907e-05],
       [-1.32435711e-04, -8.63672049e-06],
       [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Finally, using the affinity model and initialization, we can create an embedding object:
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
>>> embedding[:3]
TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
               [-1.32435711e-04, -8.63672049e-06],
               [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Notice that the embedding object contains the same point coordinates as the initialization: although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization. To optimize the embedding, we have to call the \code{embedding.optimize} method using our desired parameter settings. We use the default parameter settings here, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding.optimize(250, exaggeration=12, inplace=True)
>>> embedding.optimize(500, inplace=True)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
We can validate that method produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
0.0
\end{CodeInput}
\end{CodeChunk}
Next, we can add new samples into the embedding,
\begin{CodeChunk}
\begin{CodeInput}
>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
>>> new_embedding.optimize(
...     exaggeration=1.5, n_iter=250, learning_rate=0.1,
...     max_grad_norm=0.25, inplace=True
... )
>>> new_embedding[::3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
and validate that this also produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
0.0
\end{CodeInput}
\end{CodeChunk}

>>>>

We, again, begin by loading the Iris data set using \pkg{scikit-learn}:
\begin{CodeChunk}
\begin{CodeInput}
>>> from sklearn import datasets
>>>
>>> iris = datasets.load_iris()
>>> iris.data[:3]
array([[5.1, 3.5, 1.4, 0.2],
       [4.9, 3. , 1.4, 0.2],
       [4.7, 3.2, 1.3, 0.2]])
\end{CodeInput}
\end{CodeChunk}
Next, we import the \opentsne library:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
\end{CodeInput}
\end{CodeChunk}
We now mirror the three t-SNE steps from the previous section. First, we select an affinity model which describes the similarities between data points:
\begin{CodeChunk}
\begin{CodeInput}
>>> affinities = openTSNE.affinity.PerplexityBasedNN(
...     iris.data, perplexity=30
... )
\end{CodeInput}
\end{CodeChunk}
Next, we determine the initial point positions in the embedding:
\begin{CodeChunk}
\begin{CodeInput}
>>> initialization = openTSNE.initialization.pca(iris.data)
>>> initialization[:3]
array([[-1.30971087e-04,  1.55848907e-05],
       [-1.32435711e-04, -8.63672049e-06],
       [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Finally, using the affinity model and initialization, we can create an embedding object:
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding = openTSNE.TSNEEmbedding(initialization, affinities)
>>> embedding[:3]
TSNEEmbedding([[-1.30971087e-04,  1.55848907e-05],
               [-1.32435711e-04, -8.63672049e-06],
               [-1.40967409e-04, -7.07276279e-06]])
\end{CodeInput}
\end{CodeChunk}
Notice that the embedding object contains the same point coordinates as the initialization: although we have created an \code{openTSNE.TSNEEmbedding} object, we have not yet run any optimization. To optimize the embedding, we have to call the \code{embedding.optimize} method using our desired parameter settings. We use the default parameter settings here, including a shorter early-exaggeration phase, followed by a longer phase with no exaggeration.
\begin{CodeChunk}
\begin{CodeInput}
>>> embedding.optimize(250, exaggeration=12, inplace=True)
>>> embedding.optimize(500, inplace=True)
>>> embedding[:3]
TSNEEmbedding([[-0.55581917, 17.56162344],
               [-1.76742536, 20.07763453],
               [-0.58444608, 20.00052495]])
\end{CodeInput}
\end{CodeChunk}
We can validate that method produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(openTSNE.TSNE().fit(iris.data) - embedding)
0.0
\end{CodeInput}
\end{CodeChunk}
Next, we can add new samples into the embedding,
\begin{CodeChunk}
\begin{CodeInput}
>>> new_embedding = embedding.prepare_partial(iris.data[::3], perplexity=5)
>>> new_embedding.optimize(
...     exaggeration=1.5, n_iter=250, learning_rate=0.1,
...     max_grad_norm=0.25, inplace=True
... )
>>> new_embedding[::3]
PartialTSNEEmbedding([[-0.02637324, 15.69956674],
                      [-2.84200792, 16.15478451],
                      [-2.4594136 , 15.37910876]])
\end{CodeInput}
\end{CodeChunk}
We here use the \code{embedding.prepare_partial} method to determine initial positions for the new data points, then call the \code{new_embedding.optimize} method to run the optimization process. Again, we can validate that this also produces the exact same embedding as before:
\begin{CodeChunk}
\begin{CodeInput}
>>> np.linalg.norm(new_embedding - embedding.transform(iris.data[::3]))
0.0
\end{CodeInput}
\end{CodeChunk}
